Pre-Main-Sequence M-Dwarfs With Complex Quasiperiodic Variability From Four Years of TESS

### 2023-06-22 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are pre-main-sequence
  M-dwarfs with almost-periodic optical modulation that includes both
  sharp and broad troughs.  The troughs are usually superposed over
  smooth, spot-like modulation.  These objects are typically 10-100
  million years (Myr) old;  they do not have primordial gas-rich
  disks.  Nonetheless, the preferred explanations for their optical
  variability invoke long-lived material in orbit at the Keplerian
  corotation radius, for instance transiting dust clumps or stellar
  prominences.  Here, we report new CQVs discovered in TESS data
  collected between July~2018 and Sep~2022. 
%  Our search used two
%  methods, the first based on counting dips in phase-folded light
%  curves, and the second based on counting peaks in the Fourier
%  domain. 
   From our analysis of \nstarssearched\ M-dwarfs with
  $T$$<$16 and $d$$<$150\,pc, we find \ngoods\ gold-standard CQVs.
  Most of these discoveries are new, and they include the brightest
  and closest examples of this class of object known ($T$$\approx$11;
  $d$$\approx$20\,pc).  A few objects are outliers among outliers; LP
  12-502 for instance shows a ``dip complex'' with a total duration that is
  nearly constant over more than 1000 cycles, 
  but exhibits anywhere from four to eight local minima per cycle.  
  LP-502 also at
  times showed drastic shape changes over less than one cycle.
  More generally, we find that none of the CQVs maintain a fixed light
  curve shape over timescales of more than a few hundred cycles --
  they are {\it quasi}periodic.  Although the physical explanation
  behind these objects remains enigmatic, our sample may help clarify
  their origin.  In particular, we highlight the plausibility of a disintegrating
  planetesimal swarm model.
  Connections to star, disk, and exoplanet evolution are emphasized.
\end{abstract}

@@

The issue of what causes the variability that defines the CQVs has yet
to be resolved.  There are a few competing explanations.  All young
M-dwarfs are spotted, which produces flux variations over
characteristic timescales of the rotation period, $P_{\rm rot}$, and
its half-harmonic, $0.5\,P_{\rm rot}$.  The observed dips occur over
durations as short as $0.05\,P_{\rm rot}$; a ``starspot-only''
scenario is therefore highly unlikely for any of the objects with
sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The more likely
scenarios instead invoke sharp geometries with material extrinsic to
the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  In the
``clump'' scenario, opaque dust clumps orbiting near the Keplerian
corotation radius eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  In the
``prominence'' scenario, long-lived condensations of plasma are
trapped along the star’s magnetic field lines, also near corotation
\citep{2022MNRAS.514.5465W}.  And in a ``screen'' scenario, the inner
wall of a quiescent circumstellar disk blocks a portion of the stellar
surface to produce sudden dips whenever spots come into view
\citep{2019ApJ...876..127Z}.  In our view, the arguments seem best
established for the dust clump and prominence hypotheses
\citep{2023MNRAS.518.4734S,2022MNRAS.514.5465W}.  However, unambiguous
evidence has yet to be presented to demonstrate which, if any, of
these scenarios is correct.

@@

@@

We report on the existence of \ngoods\ objects that showed bonafide
CQV behavior over at least one TESS sector between July~2018 and Sep~2022.
Our search for these objects was restricted to dwarf stars bright enough
for 1\% precision photometry with TESS over 1 hour ($T$$<$16),
that were red ($\bprp<1.5$) and within 150\,pc of the Sun.
These objects are listed in Table~\ref{tab:thetable}.
This table also includes an additional \nmaybes\ candidate CQVs, whose
designation is less certain.

Analyzing the light curves of these objects, we can make the following conclusions.

\begin{itemize}
	%
	\item CQVs are {\it quasi}periodic.
	%
	\item The CQV phenomenon persists for at least 150\,Myr, based on the
	existence of CQVs in AB Dor, the Pleiades, and Psc-Eri.
	%
	\item CQVs evolve over timescales that are both secular ($>$100\,cycles) and impulsive ($<$1 cycle).
	%
	\item Individual CQVs can show multiple, slightly different periods.  LP 12-502
	is one example; others are included in Table~\ref{tab:thetable}.
	%
	\item The population-averaged duty cycle for CQVs seems to be $\approx 75$\%, based on 
	the fraction of bonafide CQVs that ``turned off'' during TESS re-observations.
\end{itemize}

@@

\begin{figure*}[!t]
	\begin{center}
    \centering
    \includegraphics[width=0.45\textwidth]{ANNOTATED_resid_TIC_402980664_P18.5611_2min_phase_timegroups.png}
		\vspace{-0.45cm}
		\caption{
      {\bf Alternative view of the evolution of LP 12-502}
      (Figure~\ref{fig:lp}), arranged to emphasize changes in transit
      times.  There are 200 binned black points per cycle; a two-harmonic
      sinusoid has been subtracted over specific chunks in time ({\bf see text}).
      Vertical gray lines are underplotted to help guide the eye to instances
      in which preferred dip phases synchronize over long baselines.
      The orange and green lines guide the eye to where dips
      appear to change the positions of their local minima.
		}
		\label{fig:lp2}
	\end{center}
\end{figure*}

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{TIC402980664_river_gist_stern_0_64_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
			\includegraphics[width=0.43\textwidth]{TIC402980664_river_gist_stern_248_315_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
		}
	\vspace{0.4cm}
			\subfloat{
		\includegraphics[width=0.43\textwidth]{TIC402980664_river_gist_stern_1411_1481_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
				\includegraphics[width=0.43\textwidth]{TIC402980664_river_gist_stern_1233_1265_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
	}
	\end{center}
	\caption{
    {\bf River plots of LP 12-502}, showing (clockwise from top-left) Sectors 18-19, 25-26, 53, and 58-59.
    A two-harmonic
    sinusoid has been subtracted over specific chunks in time ({\bf see text}).
	}
	\label{fig:lpriver0}
\end{figure*}

>>>>

% 250 word max!
\begin{abstract}
  Complex quasiperiodic variables (CQVs) are pre-main-sequence
  M-dwarfs with almost-periodic optical modulation that includes both
  sharp and broad troughs.  The troughs are usually superposed over
  smooth, spot-like modulation.  These objects are typically 10-100
  million years (Myr) old;  they do not have primordial gas-rich
  disks.  Nonetheless, the preferred explanations for their optical
  variability invoke long-lived material in orbit at the Keplerian
  corotation radius, for instance transiting dust clumps or stellar
  prominences.  Here, we report new CQVs discovered in TESS data
  collected between July~2018 and Sep~2022.  From our analysis of
  \nstarssearched\ M-dwarfs with $T$$<$16 and $d$$<$150\,pc, we find
  \ngoods\ gold-standard CQVs.  Most of these discoveries are new, and
  they include the brightest and closest examples of this class of
  object known ($T$$\approx$11; $d$$\approx$20\,pc).  A few objects
  are outliers among outliers; LP 12-502 for instance shows a ``dip
  complex'' with a total duration that is nearly constant over more
  than 1000 cycles, but which exhibits anywhere from four to eight
  local minima per cycle.  LP-502 also demonstrated drastic shape
  changes over less than one cycle, and has up to four distinct
  periods.  More generally, we find that none of the CQVs maintain a
  fixed light curve shape over timescales of more than a few hundred
  cycles -- they are all {\it quasi}periodic.  Although the physical
  explanation behind these objects remains enigmatic, our sample may
  help clarify their origin.  Possible connections to star, disk, and
  exoplanet evolution are emphasized.
\end{abstract}

@@

%TODO rewrite, but favor dust clumps even more
The issue of what causes the variability that defines the CQVs has yet
to be resolved.  There are a few competing explanations.  All young
M-dwarfs are spotted, which produces flux variations over
characteristic timescales of the rotation period, $P_{\rm rot}$, and
its half-harmonic, $0.5\,P_{\rm rot}$.  The observed dips occur over
durations as short as $0.05\,P_{\rm rot}$; a ``starspot-only''
scenario is therefore highly unlikely for any of the objects with
sufficiently sharp dips
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The more likely
scenarios instead invoke sharp geometries with material extrinsic to
the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  In the
``clump'' scenario, opaque dust clumps orbiting near the Keplerian
corotation radius eclipse the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  In the
``prominence'' scenario, long-lived condensations of plasma are
trapped along the star’s magnetic field lines, also near corotation
\citep{2022MNRAS.514.5465W}.  And in a ``screen'' scenario, the inner
wall of a quiescent circumstellar disk blocks a portion of the stellar
surface to produce sudden dips whenever spots come into view
\citep{2019ApJ...876..127Z}.  In our view, the arguments seem best
established for the dust clump and prominence hypotheses
\citep{2023MNRAS.518.4734S,2022MNRAS.514.5465W}.  However, unambiguous
evidence has yet to be presented to demonstrate which, if any, of
these scenarios is correct.

@@

{\sc Dip periods}---Most of the LP 12-502 dips recur with a period of
$P=18.5611 \pm 0.0001$\,hr.  However the river plots reveal multiple
distinct periodicities in the light curve for specific dips.  For
instance, in sectors 25-26, the local minimum that develops around
cycle 262 has a period faster than the mean period by $\approx$0.1\%,
while some of the trailing local minima in the main dip complex have
periods slower than the mean period, by $\approx$0.04\%.  In addition
to the fundamental period, we were able to identify at least four
distinct periods shown by specific dips over the full Sectors 18-59
dataset, including periods at 18.5683, 18.5672, 18.5473, and
18.5145\,hr, with a typical measurement uncertainty of
$\approx$0.0002\,hr.

@@

We report on the existence of \ngoods\ objects that showed
complex quasiperiodic behavior over at least one TESS sector between July~2018 and Sep~2022.
Our search for these objects was restricted to dwarf stars bright enough
for 1\% precision photometry with TESS over 1 hour ($T$$<$16),
that were red ($\bprp<1.5$) and within 150\,pc of the Sun.
These \ngoods\ bonafide CQVs are listed in Table~\ref{tab:thetable}.
This table also includes an additional \nmaybes\ candidate CQVs, whose
designation is less certain.

Analyzing the light curves of these objects, we draw the following conclusions.

\begin{itemize}
	%
	\item CQVs are {\it quasi}periodic.  While the mean periods in our
	sample seem to remain fixed over the span of available observations, the light curve
	shapes themselves evolve, similar to rotating stars.
	%
	\item The same CQV can show dips with closely-spaced periods.  
	LP 12-502, for instance, showed dips with four distinct periods within $\pm 0.3\%$
	of the fundamental period, sometimes simultaneous, and each lasting for up to 50 cycles.
	%
	\item The CQV phenomenon persists for at least 150\,Myr, based on the
	existence of CQVs in AB Dor, the Pleiades, and Psc-Eri.  It may extend to
	even older ages, however the lack of detected CQVs in the Hyades and
	Praesepe suggest that the phenomenon is unique to the pre-main-sequence
	for M-dwarfs.
	%
	\item CQVs evolve over timescales that are both secular ($>$100\,cycles) and impulsive ($<$1 cycle).
	``State-switches'' can cause dips to collapse instantaneously, and are often (but not exclusively)
	linked with observed optical flares.
	Dip growth however seems to happen over durations of at least ten cycles.
	%
	\item The population-averaged duty cycle for CQVs seems to be $\approx 75$\%, based on 
	the fraction of bonafide CQVs that ``turned off'' during TESS re-observations.
\end{itemize}

@@

\begin{figure*}[!t]
	\begin{center}
    \includegraphics[width=0.45\textwidth]{ANNOTATED_resid_TIC_402980664_P18.5611_2min_phase_timegroups.png}
    	\end{center}
    \vspace{-0.4cm}
		\caption{
	      {\bf Alternative view of the evolution of LP 12-502}
	      (Figure~\ref{fig:lp}), arranged to emphasize changes in transit
	      times.  There are 200 binned black points per cycle; a two-harmonic
	      sinusoid has been subtracted over specific chunks in time ({\bf see text}).
	      Vertical gray lines are underplotted to help guide the eye to instances
	      in which preferred dip phases synchronize over long baselines.
	      The orange and green lines guide the eye to where dips
	      appear to change the positions of their local minima.
		}
		\label{fig:lp2}
\end{figure*}

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{TIC402980664_river_gist_stern_0_64_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
			\includegraphics[width=0.43\textwidth]{ANNOTATED_TIC402980664_river_gist_stern_248_315_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
		}
	\vspace{-0.1cm}
			\subfloat{
		\includegraphics[width=0.43\textwidth]{ANNOTATED_TIC402980664_river_gist_stern_1411_1481_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
				\includegraphics[width=0.43\textwidth]{ANNOTATED_TIC402980664_river_gist_stern_1233_1265_manual_20230617_mask_v0_nterms2_vmin-0.04000_vmax0.01000.png}
	}
	\end{center}
	\caption{
    {\bf River plots of LP 12-502}, showing (clockwise from top-left)
    Sectors 18-19, 25-26, 53, and 58-59.  A two-harmonic sinusoid has
    been subtracted over specific chunks in time ({\bf see text}).
    For Sectors 25-26 (cycles 248-315), three periods are overplotted:
    $P$=18.5611\,hr (gray vertical line); 18.5404\,hr (orange); 18.5683\,hr (green).
    For Sector 53, gray is identical, while cyan is 18.5145\,hr.
    For Sectors 58-59, the magenta line is 18.5473\,hr, and the green
    line is 18.5672\,hr.
	}
	\label{fig:lpriver0}
\end{figure*}
